---
title: "Pharmacy First Data Development Report"
output: 
  html_document:
  toc: true
  toc_depth: 4
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

This markdown report creates the tables and figures for the **Pharmacy First Data Development** project. 

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(gt)
library(webshot2)
library(janitor)
library(httr)
```

```{r load-data, include=FALSE}
output_dir <- "released_output"
df_med_status_counts <- read_csv(
  here(output_dir, "data_development", "med_status_counts.csv")
)
df_pf_codes_counts <- read_csv(
  here(output_dir, "data_development", "pf_codes_counts.csv")
)
df_codes_count_measures <- read_csv(
  here(output_dir, "clinical_codes", "code_counts_measures.csv")
)
df_codes_count_summary <- read_csv(
  here(output_dir, "data_development", "pf_codes_count_summary.csv")
)
df_consultation_med_counts <- read_csv(
  here(output_dir, "measures", "consultation_med_counts_measures.csv"),
  col_types = cols(dmd_code = col_character())
)
```

## Methods

We utilised the OpenSAFELY framework to conduct a retrospective data development study, aiming to explore codes relevant to the Pharmacy First service.

### Codelists

To explore the codes relevant to Pharmacy First, we utilised various codelists to identify specific events and medications.


#### Pharmacy First Event Codes

The following two codes were used interchangeably to describe a Pharmacy First consultation. To identify consultations classified as 'Pharmacy First consultations,' we utilised the SNOMED codes below:

- **Community Pharmacist (CP) Consultation Service for minor illness (procedure)**: `"1577041000000109"`
- **Pharmacy First service (qualifier value)**: `"983341000000102"`


#### Pharmacy First Clinical Pathways Codes

We employed the Pharmacy First [clinical pathways codelist](https://www.opencodelists.org/codelist/opensafely/pharmacy-first-clinical-pathway-conditions/7ec97762/#full-list) to categorise clinical events related to Pharmacy First services according to specific Pharmacy First pathways.


#### Pharmacy First Medication Codes (by Clinical Pathway)

To isolate medications commonly prescribed by pharmacies for each of the seven Pharmacy First clinical pathways, we utilised seven distinct codelists containing NICE-recommended medications for each pathway. The following codelists were used to identify the supply of any medication included in the Pharmacy First Patient Group Direction for the treatment of Pharmacy First conditions:

- [Acute otitis media](https://www.opencodelists.org/codelist/opensafely/pharmacy-first-acute-otitis-media-treatment-dmd/4122468e/#full-list)
- [Impetigo](https://www.opencodelists.org/codelist/opensafely/pharmacy-first-impetigo-treatment-dmd/708e6d02/#full-list)
- [Infected insect bites](https://www.opencodelists.org/codelist/opensafely/pharmacy-first-infected-insect-bites-treatment-dmd/2812d359/#full-list)
- [Shingles](https://www.opencodelists.org/codelist/opensafely/pharmacy-first-shingles-treatment-dmd/3bbcc6a2/#full-list)
- [Sinusitis](https://www.opencodelists.org/codelist/opensafely/pharmacy-first-sinusitis-treatment-dmd/0f036024/#full-list)
- [Sore throat](https://www.opencodelists.org/codelist/opensafely/pharmacy-first-sore-throat-treatment-dmd/2ac5934f/#full-list)
- [Urinary tract infection](https://www.opencodelists.org/codelist/opensafely/pharmacy-first-urinary-tract-infection-treatment-dmd/54cc39d7/#full-list)

```{r tidy-data, include=FALSE}
# Medication status labels and descriptions
med_status_label <- paste0("status", 0:28)

med_status_label_desc <- c(
  "0 - Normal",
  NA_character_, NA_character_, NA_character_,
  "4 - Historical",
  "5 - Blue script",
  "6 - Private",
  "7 - Not in possession",
  "8 - Repeat dispensed",
  "9 - In possession",
  "10 - Dental",
  "11 - Hospital",
  "12 - Problem substance",
  "13 - From patient group direction",
  "14 - To take out",
  "15 - On admission",
  "16 - Regular medication",
  "17 - As required medication",
  "18 - Variable dose medication",
  "19 - Rate-controlled single regular",
  "20 - Only once",
  "21 - Outpatient",
  "22 - Rate-controlled multiple regular",
  "23 - Rate-controlled multiple only once",
  "24 - Rate-controlled single only once",
  "25 - Placeholder",
  "26 - Unconfirmed",
  "27 - Infusion",
  "28 - Reducing dose blue script"
)

# Pharmacy first codes
pf_consultation_label <- c(
  "count_consultation_service",
  "count_pharmacy_first_service"
)

pf_consultation_label_desc <- c(
  "Consultation Service for minor illness (1577041000000109)",
  "Pharmacy First service (983341000000102)"
)

summary_stat_label <- c(
  "count_distinct_ids",
  "count_pf_events",
  "count_pf_pathways"
)

summary_stat_label_desc <- c(
  "Count of all events",
  "Count of PF events",
  "Count of all PF pathways"
)
```

## Medication status

Initially, we investigated the [`medication_status`](https://docs.opensafely.org/ehrql/reference/schemas/raw.tpp/#medications.medication_status) field within the [`medications`](https://docs.opensafely.org/ehrql/reference/schemas/raw.tpp/#medications) table by examining all possible values with counts over a one-year period. 
The table below illustrates this, alongside a breakdown of all medications in the dataset by their respective status. Subsequently, the Pharmacy First medication status counts are further disaggregated by splitting the one-year period into two six-month intervals: the six months prior to the launch of Pharmacy First, and the six months following it.

Further analysis was conducted by categorising pre- and post-launch data into three groups: all medications, exclusively Pharmacy First medications, and instances where Pharmacy First medications were issued on the same day as a Pharmacy First consultation. This approach helps evaluate the potential impact of the Pharmacy First launch on medication status patterns.

### Overall medication status counts

```{r, message=FALSE, warning=FALSE, echo = FALSE}
df_tab_med_status_counts <- df_med_status_counts %>%
  replace_na(list(n = 0)) %>%
  filter(selected_codes == "anymed") %>%
  group_by(med_status) %>%
  mutate(
    n = sum(n, na.rm = TRUE),
    med_status = factor(
      med_status,
      levels = med_status_label,
      labels = med_status_label_desc
    )
  ) %>%
  select(-time, -selected_codes) %>%
  distinct() %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

tab_med_status_counts <- df_tab_med_status_counts %>%
  gt() %>%
  sub_missing(
    columns = everything(),
    rows = everything(),
    missing_text = "---"
  ) %>%
  fmt_number(
    columns = n,
    decimals = 0
  ) %>%
  fmt_percent(
    columns = pct,
    decimals = 3
  ) %>%
  cols_label(
    med_status = "Medication status",
    n = "Count",
    pct = "Percent"
  ) %>%
  cols_align(
    align = "left",
    columns = med_status
  ) %>%
  tab_header(
    title = md("**Descriptive statistics for medication status**"),
    subtitle = "This includes all medications issued between 1st Aug 2023 and 31st Jul 2024",
  )

gtsave(
  tab_med_status_counts,
  here("reports", "resources", "tab_med_status_counts.png"),
)

tab_med_status_counts
```

### Pharmacy First medication status counts

```{r, message=FALSE, warning=FALSE, echo = FALSE}
# Process data frame
df_tab_pf_med_status_counts <- df_med_status_counts %>%
  replace_na(list(n = 0)) %>%
  pivot_wider(names_from = c(time, selected_codes), values_from = n) %>%
  mutate(
    med_status = factor(
      med_status,
      levels = med_status_label,
      labels = med_status_label_desc
    )
  ) %>%
  select(
    med_status,
    post_anymed, post_anypfid,
    post_pfmed, post_pfmedid,
  )

# Display data frame as table
tab_pf_med_status_counts <- df_tab_pf_med_status_counts %>%
  filter(med_status %in% c("0 - Normal", "4 - Historical", "8 - Repeat dispensed", "11 - Hospital")) %>%
  gt() %>%
  sub_missing(
    columns = everything(),
    rows = everything(),
    missing_text = "---"
  ) %>%
  fmt_number(
    columns = c(
      post_anymed,
      post_anypfid,
      post_pfmed,
      post_pfmedid,
    ),
    decimals = 0
  ) %>%
  tab_spanner(
    label = md("**Any medication**"),
    columns = contains("any")
  ) %>%
  tab_spanner(
    label = md("**Pharmacy First medications**"),
    columns = contains("pfmed")
  ) %>%
  cols_label(
    med_status = md("**Medication status**"),
    post_anymed = "All",
    post_anypfid = "PF ID",
    post_pfmed = "All",
    post_pfmedid = "PF ID",
  ) %>%
  cols_align(
    align = "left",
    columns = med_status
  ) %>%
  tab_footnote(
    footnote = "All medications.",
    locations = cells_column_labels(columns = c(post_anymed, post_pfmed))
  ) %>%
  tab_footnote(
    footnote = "Medications with Pharmacy First consultation ID.",
    locations = cells_column_labels(columns = c(post_anypfid, post_pfmedid))
  ) %>%
  tab_header(
    title = md("**Breakdown of medication status for Pharmacy First service**"),
    subtitle = "Timeframe 1st Feb 2024 to 31st July 2024",
  )

gtsave(
  tab_pf_med_status_counts,
  here("reports", "resources", "tab_med_status_counts.png"),
)

tab_pf_med_status_counts
```

### Medication count validation 
```{r, message=FALSE, warning=FALSE, echo = FALSE}
df_med_count_validation <- read_csv(
  (here("lib", "validation", "data", "eps_erd_prescribing_2024-02-01_to_2024-07-01.csv"))
)

tab_med_count_validation  <- df_med_count_validation %>%
  pivot_longer(
    cols = c(eps_items, erd_items),
    names_to = "med_status",
    values_to = "count"
  ) %>%
  mutate(count_40pct = round(count * .4, digits = 0)) %>%
  group_by(med_status) %>%
  summarise(post_nhsbsa_40pct = sum(count_40pct, na.rm = TRUE)) %>%
  mutate(med_status = factor(med_status, levels = c("eps_items", "erd_items"), labels = c("0 - Normal", "8 - Repeat dispensed")))

```
In order to validate that the number of medications that we see is as we expect, we will use the publicly available [Electronic Prescription Service (EPS) and electronic Repeat Dispensing (eRD) data](https://www.nhsbsa.nhs.uk/access-our-data-products/epact2/dashboards-and-specifications/electronic-prescription-service-eps-and-electronic-repeat-dispensing-erd-utilisation-dashboard) published by the NHSBSA as validation data. 
The EPS and eRD data relate to the '0 - Normal' and '8 - Repeat dispensed' medication statuses from the table above. 

We expected to see roughly 40% of this overall data for the months we included in our study, as shown in the table below:

```{r, message=FALSE, warning=FALSE, echo = FALSE}
# Add new column which contains validation values
df_filtered_med_status_counts <- df_tab_pf_med_status_counts %>%
  filter(med_status %in% c("0 - Normal", "8 - Repeat dispensed")) %>%
  left_join(tab_med_count_validation)

tab_any_validation <- df_filtered_med_status_counts %>%
  select(med_status, post_anymed, post_nhsbsa_40pct) %>%
  gt() %>%
  tab_header(
    title = "Comparison of EPS and ERD values to public (validation) data",
    subtitle = "Timeframe 1st Feb 2024 to 31st July 2024",
  ) %>%
  cols_label(
    med_status = "Med Status",
    post_anymed = "Any Medication",
    post_nhsbsa_40pct = "Medication (Validation)"
  ) %>%
  fmt_number(
    columns = vars(post_anymed, post_nhsbsa_40pct),
    decimals = 0
  ) %>%
  tab_footnote(
    footnote = "40% of data taken from NHSBSA EPS and eRD dispensing data - July 2024",
    locations = cells_column_labels(columns = c(post_nhsbsa_40pct))
  )
# Display the table for validation of 'any medication'
tab_any_validation
```