---
title: "Pharmacy First Data Development Report"
output: 
  html_document:
  toc: true
  toc_depth: 4
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

This markdown report creates the tables and figures for the **Pharmacy First Data Development** project. 

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(gt)
library(webshot2)
```

```{r load-data, include=FALSE}
output_dir <- "released_output"
df_med_status_counts <- read_csv(
  here(output_dir, "data_development", "med_status_counts.csv")
)
df_pf_codes_counts <- read_csv(
  here(output_dir, "data_development", "pf_codes_counts.csv")
)
df_codes_count_measures <- read_csv(
  here(output_dir, "clinical_codes", "code_counts_measures.csv")
)
df_codes_count_summary <- read_csv(
  here(output_dir, "data_development", "pf_codes_count_summary.csv")
)
```

## Methods

We utilised the OpenSAFELY framework to conduct a retrospective data development study, aiming to explore codes relevant to the Pharmacy First service.

```{r tidy-data, include=FALSE}
# Medication status labels and descriptions
med_status_label <- paste0("status", 0:28)

med_status_label_desc <- c(
  "0 - Normal",
  NA_character_, NA_character_, NA_character_,
  "4 - Historical",
  "5 - Blue script",
  "6 - Private",
  "7 - Not in possession",
  "8 - Repeat dispensed",
  "9 - In possession",
  "10 - Dental",
  "11 - Hospital",
  "12 - Problem substance",
  "13 - From patient group direction",
  "14 - To take out",
  "15 - On admission",
  "16 - Regular medication",
  "17 - As required medication",
  "18 - Variable dose medication",
  "19 - Rate-controlled single regular",
  "20 - Only once",
  "21 - Outpatient",
  "22 - Rate-controlled multiple regular",
  "23 - Rate-controlled multiple only once",
  "24 - Rate-controlled single only once",
  "25 - Placeholder",
  "26 - Unconfirmed",
  "27 - Infusion",
  "28 - Reducing dose blue script"
)

# Pharmacy first codes
pf_consultation_label <- c(
  "count_consultation_service",
  "count_pharmacy_first_service"
)

pf_consultation_label_desc <- c(
  "Consultation Service for minor illness (1577041000000109)",
  "Pharmacy First service (983341000000102)"
)

summary_stat_label <- c(
  "count_distinct_ids",
  "count_pf_events",
  "count_pf_pathways"
)

summary_stat_label_desc <- c(
  "Count of all events",
  "Count of PF events",
  "Count of all PF pathways"
)
```

## Medication status

Initially, we investigated the [`medication_status`](https://docs.opensafely.org/ehrql/reference/schemas/raw.tpp/#medications.medication_status) field within the [`medications`](https://docs.opensafely.org/ehrql/reference/schemas/raw.tpp/#medications) table by examining all possible values with counts over a one-year period. 
The table below illustrates this, alongside a breakdown of all medications in the dataset by their respective status. Subsequently, the Pharmacy First medication status counts are further disaggregated by splitting the one-year period into two six-month intervals: the six months prior to the launch of Pharmacy First, and the six months following it.

Further analysis was conducted by categorising pre- and post-launch data into three groups: all medications, exclusively Pharmacy First medications, and instances where Pharmacy First medications were issued on the same day as a Pharmacy First consultation. This approach helps evaluate the potential impact of the Pharmacy First launch on medication status patterns.

### Overall medication status counts

```{r, message=FALSE, warning=FALSE, echo = FALSE}
df_tab_med_status_counts <- df_med_status_counts %>%
  replace_na(list(n = 0)) %>%
  filter(selected_codes == "any") %>%
  group_by(med_status) %>%
  mutate(
    n = sum(n, na.rm = TRUE),
    med_status = factor(
      med_status,
      levels = med_status_label,
      labels = med_status_label_desc
    )
  ) %>%
  select(-time, -selected_codes) %>%
  distinct() %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

tab_med_status_counts <- df_tab_med_status_counts %>%
  gt() %>%
  sub_missing(
    columns = everything(),
    rows = everything(),
    missing_text = "---"
  ) %>%
  fmt_number(
    columns = n,
    decimals = 0
  ) %>%
  fmt_percent(
    columns = pct,
    decimals = 3
  ) %>%
  cols_label(
    med_status = "Medication status",
    n = "Count",
    pct = "Percent"
  ) %>%
  cols_align(
    align = "left",
    columns = med_status
  ) %>%
  tab_header(
    title = md("**Descriptive statistics for medication status**"),
    subtitle = "This includes all medications issued between 1st Aug 2023 and 31st Jul 2024",
  )

gtsave(
  tab_med_status_counts,
  here("reports", "resources", "tab_med_status_counts.png"),
)

tab_med_status_counts
```

### Pharmacy First medication status counts

```{r, message=FALSE, warning=FALSE, echo = FALSE}
# Process data frame
df_tab_pf_med_status_counts <- df_med_status_counts %>%
  replace_na(list(n = 0)) %>%
  pivot_wider(names_from = c(time, selected_codes), values_from = n) %>%
  mutate(
    pre_pfmedid = NA,
    med_status = factor(
      med_status,
      levels = med_status_label,
      labels = med_status_label_desc
    )
  ) %>%
  relocate(
    med_status,
    pre_any, pre_pfmed, pre_pfmedid,
    post_any, post_pfmed, post_pfmedid
  )

# Display data frame as table
tab_pf_med_status_counts <- df_tab_pf_med_status_counts %>%
  gt() %>%
  sub_missing(
    columns = everything(),
    rows = everything(),
    missing_text = "---"
  ) %>%
  fmt_number(
    columns = c(
      pre_any,
      pre_pfmed,
      pre_pfmedid,
      post_any,
      post_pfmed,
      post_pfmedid
    ),
    decimals = 0
  ) %>%
  tab_spanner(
    label = "6 months pre PF launch",
    columns = starts_with("pre_")
  ) %>%
  tab_spanner(
    label = "6 months post PF launch",
    columns = starts_with("post_")
  ) %>%
  cols_label(
    med_status = "Medication status",
    pre_any = "Any",
    pre_pfmed = "PF Med",
    pre_pfmedid = "PF Med & Consult",
    post_any = "Any",
    post_pfmed = "PF Med",
    post_pfmedid = "PF Med & Consult"
  ) %>%
  cols_align(
    align = "left",
    columns = med_status
  ) %>%
  tab_footnote(
    footnote = "Any medication.",
    locations = cells_column_labels(columns = c(pre_any, post_any))
  ) %>%
  tab_footnote(
    footnote = "Pharmacy First medications.",
    locations = cells_column_labels(columns = c(pre_pfmed, post_pfmed))
  ) %>%
  tab_footnote(
    footnote = "Pharmacy First medications and Pharmacy First consultation.",
    locations = cells_column_labels(columns = c(pre_pfmedid, post_pfmedid))
  ) %>%
  tab_header(
    title = md("**Descriptive statistics for medication status**"),
    subtitle = "Counts are broken down by type of medication and timeframe",
  )

gtsave(
  tab_pf_med_status_counts,
  here("reports", "resources", "tab_med_status_counts.png"),
)

tab_pf_med_status_counts
```

## Pharmacy First consultation codes

We extracted the consultation IDs associated with consultations that used pharmacy first event SNOMED codes from the clinical events dataset.
We then visualised the total counts of consultation IDs linked to Pharmacy First, followed by a breakdown across individual clinical services to assess changes in consultation frequency after the Pharmacy First launch. The primary objective of this analysis was to establish evidence that consultations linked with Pharmacy First events could reliably serve as 'Pharmacy First consultations' in future studies.

To provide further insight, consultation ID counts were segmented by clinical service. As these services were in use prior to the Pharmacy First launch, examining background levels and regional patterns allows us to assess their pre-launch usage and identify any post-launch increase, acknowledging that different regions may exhibit varied coding behaviours.

```{r, fig.height=5, fig.width=8, echo = FALSE, warning=FALSE}
# Read the CSV file
plot_code_counts <- df_codes_count_measures %>%
  group_by(measure, interval_end) %>%
  mutate(
    numerator_total = sum(numerator, na.rm = TRUE),
    denominator_total = sum(denominator, na.rm = TRUE),
    measure = factor(measure, levels = pf_consultation_label, labels = pf_consultation_label_desc),
  ) %>%
  select(measure, interval_end, numerator_total, denominator_total) %>%
  filter(measure %in% pf_consultation_label_desc) %>%
  distinct() %>%
  ggplot(aes(
    x = interval_end,
    y = numerator_total,
    colour = measure,
    shape = measure,
  )) +
  geom_point(size = 2) +
  geom_line(alpha = .3) +
  labs(
    title = NULL,
    x = NULL,
    y = "Total count",
    colour = NULL,
    shape = NULL
  ) +
  scale_y_continuous(
    labels = scales::label_number(),
  ) +
  theme(legend.position = "bottom") +
  guides(
    colour = guide_legend(ncol = 2),
    shape = guide_legend(ncol = 2)
  ) +
  scale_x_date(
    date_breaks = "1 month",
    labels = scales::label_date_short()
  ) +
  geom_vline(
    xintercept = lubridate::as_date(c(
      "2024-01-31"
    )),
    linetype = "dotted",
    colour = "orange",
    size = .7
  ) +
  scale_colour_viridis_d(end = .75)

ggsave(
  here("reports", "resources", "plot_code_counts.png"),
  plot_code_counts,
  height = 5,
  width = 8
)

plot_code_counts
```

```{r, fig.height=10, fig.width=8, echo = FALSE, warning=FALSE}
# Read the CSV file

# Process data frame
plot_code_counts_region <- df_codes_count_measures %>%
  replace_na(list(practice_region = "(Missing)")) %>%
  mutate(
    measure = factor(measure, levels = pf_consultation_label, labels = pf_consultation_label_desc),
    practice_region = fct_relevel(practice_region, "(Missing)", after = Inf)
  ) %>%
  filter(measure %in% pf_consultation_label_desc) %>%
  ggplot(aes(
    x = interval_end,
    y = ratio,
    colour = practice_region,
  )) +
  geom_point() +
  geom_line(alpha = .3) +
  labs(
    title = NULL,
    x = NULL,
    y = "Ratio of counts",
    colour = NULL
  ) +
  scale_y_continuous(
    labels = scales::label_percent(),
  ) +
  facet_wrap(~measure,
    scales = "free_y",
    ncol = 1
  ) +
  scale_colour_brewer(palette = "Set1", na.value = "#000000") +
  theme(legend.position = "bottom") +
  scale_x_date(
    date_breaks = "1 month",
      labels = scales::label_date_short()
    ) +
    geom_vline(
      xintercept = lubridate::as_date(c(
        "2024-01-31"
      )),
      linetype = "dotted",
      colour = "orange",
      size = .7
    )

ggsave(
  here("reports", "resources", "plot_code_counts_region.png"),
  plot_code_counts_region,
  height = 12,
  width = 8
)

plot_code_counts_region
```

### Pharmacy First counts summary

The table below summarises the total count of clinical events (sourced from the clinical events dataset) before and after the Pharmacy First launch. This summary provides a visual representation of the rise in Pharmacy First events since the launch, as well as clinical events concerning conditions linked to Pharmacy First pathways. Additionally, further breakdowns include counts of clinical events occurring on the same day as a Pharmacy First consultation, and those with a Pharmacy First consultation ID, to identify potential instances of missing Pharmacy First clinical events. 

```{r, message=FALSE, warning=FALSE, echo = FALSE}
df_pivot_codes_count_summary <- df_codes_count_summary %>%
  pivot_wider(names_from = c(time, selected_events), values_from = count) %>%
  mutate(
    summary_stat = factor(
      summary_stat,
      levels = summary_stat_label,
      labels = summary_stat_label_desc
    )
  ) %>%
  relocate(summary_stat,
    pre_all, pre_pfdate, pre_pfid,
    post_all, post_pfdate, post_pfid
  )

# Display data frame as table
tab_pf_codes_count_summary <- df_pivot_codes_count_summary %>%
  gt() %>%
  sub_missing(
    columns = everything(),
    rows = everything(),
    missing_text = "---"
  ) %>%
  fmt_number(
    columns = c(
      pre_all,
      pre_pfdate,
      pre_pfid,
      post_all,
      post_pfdate,
      post_pfid
    ),
    decimals = 0
  ) %>%
  tab_spanner(
    label = "6 months pre PF launch",
    columns = starts_with("pre_")
  ) %>%
  tab_spanner(
    label = "6 months post PF launch",
    columns = starts_with("post_")
  ) %>%
  cols_label(
    summary_stat = "Summary Statistic",
    pre_all = "Any event",
    pre_pfdate = "Same day as PF event",
    pre_pfid = "Same ID as PF event",
    post_all = "Any event",
    post_pfdate = "Same day as PF event",
    post_pfid = "Same ID as PF event"
  ) %>%
  cols_align(
    align = "left",
    columns = summary_stat
  ) %>%
  tab_footnote(
    footnote = "Total number of distinct consultation IDs.",
    locations = cells_column_labels(columns = c(pre_all, post_all))
  ) %>%
  tab_footnote(
    footnote = "Clinical events taking place on the same day as a Pharmacy First consultation.",
    locations = cells_column_labels(columns = c(pre_pfdate, post_pfdate))
  ) %>%
  tab_footnote(
    footnote = "Clinical events with a Pharmacy First consultation ID.",
    locations = cells_column_labels(columns = c(pre_pfid, post_pfid))
  ) %>%
  tab_header(
    title = md("**Descriptive statistics for clinical events related to Pharmacy First services**"),
    subtitle = "Counts are broken down by different summary statistics and timeframe",
  )

gtsave(
  tab_pf_codes_count_summary,
  here("reports", "resources", "tab_pf_codes_count_summary.png"),
)

tab_pf_codes_count_summary
```

* Pharmacy First pathways (sinusitis, sore throat, otitis media, infected insect bites, impetigo, shingles, uncomplicated urinary tract infections)